From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-08-21 WOLFHPC15/article/component.tex
Rel S5

%----------------------------------------
%\subsection{Overview}
%----------------------------------------
This section describes the projection of a MSP program to a component-based runtime, including the components used to implement the static scheduling of $\Gamma_{tsp}$.
Before those details, the section gives an overview of component models, and especially the Low Level Components (\llc) used in this paper.

@@

Many component models exist, each of them with its own specificities.
Well known component models include, for example, the CORBA Component Model (CCM)~\cite{corba:omg06}, and the Grid Component Model (GCM)~\cite{Baude} for distributed computing, while the Common Component Architecture (CCA)~\cite{Bernholdt01052006}, and Low Level Components (\llc)~\cite{l2c} are HPC-oriented.
This work makes use of \llc for the experiments.
Let us therefore introduce the concepts of this model in more details.

@@

\llc is a minimalist \texttt{C++} based HPC-oriented component model
where a component extends the concept of class.
The services offered by the components are specified trough $provide$ ports,
those used either by $use$ ports for a single service instance,
or $use-multiple$ ports for multiple service instances.
Services are specified as \texttt{C++} interfaces.
\llc also offers $MPI$ ports that enable components to share MPI communicators.
Components can also have attribute ports to be configured.
%
As illustrated in Figures~\ref{fig:ports}, a $provide$ port is
represented by a white circle, a $use$
port with a black circle, a $use-multiple$ port by a black circle with
a white $m$ in it. MPI port are
connected with a black rectangle.
A \llc-based application is a static \emph{assembly} of components instances and connections between their ports.
Such an assembly is described in LAD, an XML dialect~\cite{l2c}.

@@

%----------------------------------------
\subsection{Runtime Overview}
%----------------------------------------

@@

  \path[-]
    (Dp) edge node {} (D)
    (D) edge node {} (Du1)
        edge node {} (Du2)
        edge node {} (Du3)
    (DDSp1) edge node {} (DDS)
    (Tp) edge node {} (T)
    (T)  edge node {} (Tu)
    (Cp) edge node {} (C)
    (C) edge node {} (Cu)
    (Datap2) edge node {} (Data)
    (Datap1) edge node {} (Data)
    (Data) edge node {} (Datau)
          edge node {} (mpidata)
    (DDSp2) edge node {} (DDS);
\end{tikzpicture}
\vspace*{.5em}
\caption{MSCAC Component Assembly Back-End Overview.}
\label{fig:mscac:assembly}
\end{center}
\end{figure}

As described in Section~\ref{sect:mscac}, the compiler generates four independent parts in a component assembly in addition to a static part.
This is represented on Figure~\ref{fig:mscac:assembly} where each rounded box represents one component instance, drawn with simple lines, or more, drawn with double lines. For example, the component \texttt{Data} is instanciated more than once.
\texttt{Driver} is static and drives the whole application.
The four other parts are generated based on the content of the different sections of the MSL program: \texttt{Computations}, from the parallelization of $\Gamma$ (done by MSC as explained in Section~\ref{sect:msc}), \emph{DDS} from $\mathcal{M}$ and $\mathcal{E}$, \emph{Data} from $\Delta$ and $\mathcal{D}$, and \emph{Time} from $T$.

@@

\texttt{DDS} manages the structure and the partitioning of the mesh, a single instance is generated as MSL only supports a single mesh for now.
The back-end presented in this paper uses the distributed data structure proposed in SkelGIS library~\cite{CPE:CPE3494}.
\texttt{DDS} is used by \texttt{Data}, for which one instance represents one data element in the MSL program.
\texttt{Time} is parametrized by the the \texttt{iteration} value from MSL and uses the \texttt{Computations} component as many times as required.

\texttt{Computations} contains the static scheduling $\Gamma_{tsp}$, or the data parallel transformation $\Gamma_{data}$, both computed by MSC.
Both transformations, $\Gamma_{tsp}$ and $\Gamma_{data}$, are encoded using three specific components, detailed in Section~\ref{sect:control}, which manage the $P$, $S$, and $sync$ operations.
At the leaves of $\Gamma_{tsp}$ and $\Gamma_{data}$, the computation components instances (denoted $K$), provided by the user, are connected to \texttt{Data}.
As presented on Figure~\ref{fig:k}, a $K$ component provides a single port used to launch the computation and exposes one use port for each data element manipulated in the computation.
We denote the fact that it may have several use port by adding a star on the use port to \texttt{Data}.

\begin{figure}
\begin{center}
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[component] (k) at (0,0) {$K$};
   \node[provide] (p) at (-1.5,0) {};
   \node[use] (u) at (1.5,0) {};
   \node[right=0.2 of u] (star) {*};

  \path[-]
    (p) edge node {} (k)
    (k) edge node {} (u);
\end{tikzpicture}
\end{center}
\caption{Kernel component.}
\label{fig:k}
\end{figure}

Moreover, to finalize the data parallelization of the back-end component-based runtime, which is partially managed by \texttt{DDS} for the partitioning, and $sync$ operations in \texttt{Computations}, we have chosen to duplicate the component assembly of Figure~\ref{fig:mscac:assembly} on available processors (or cores), connecting them by MPI ports between their \texttt{Data}.

As a result of this approach, each part of the generated code is rather independent.
Changing the main time loop to use a convergence criteria rather than a fixed number of iterations would only require changing the \texttt{Time} component.
Similarly, changing the approach used for scheduling would only impact the code generated in \texttt{Computations}.
The technology used for data parallelism can also be changed by replacing the \texttt{DDS} and \texttt{Data} components; and by using the adequate interfaces in $K$. As a matter of fact, a work in progress proposes an alternative based on Global Arrays~\cite{Nieplocha:2006:AAP:1125980.1125985} instead of SkelGIS.

%----------------------------------------
\subsection{Control Components and Dump}
\label{sect:control}
%----------------------------------------
The series-parallel tree decomposition $\Gamma_{tsp}$ represents a static scheduling of the simulation. $\Gamma_{data}$, on the other hand, is an optimized fusionned ordered list of computations. Both are dumped to components by generating an assembly that exactly matches their structure.
We introduce what we call \emph{control components} to represent all nodes of $\Gamma_{tsp}$ or $\Gamma_{data}$.
These components can be used for any simulation, which increases code reuse between simulations.
A control component exposes a single \emph{provide} port containing a \emph{control-only} method, without any parameter.
We introduce three control components represented in Figure~\ref{fig:ctrlcomponents}.

@@

  \path[-]
    (p) edge node {} (sync)
    (sync) edge node {} (u);
\end{tikzpicture}
\\
a) Comp. SEQ
\hspace{\fill}
b) Comp. PAR
\hspace{\fill}
c) Comp. SYNC
\\
\caption{The three control components used by the back-end.}
\label{fig:ctrlcomponents}
\end{figure}

\item[Synchronization component (SYNC)] It is the direct representation to an update leaf of $\Gamma_{tsp}$ or $\Gamma_{data}$. This component calls the synchronization of a given data. It exposes a \emph{use} port to be connected to the data to update.
\end{description}

Then from $\Gamma_{tsp}$ or $\Gamma_{data}$, and using the control components described above,
a direct dump can be done to a component assembly for a data or hybrid (data
and task) parallelization of a simulation.
Figure~\ref{fig:tsp:assembly} displays the assembly part corresponding to
$\Gamma_{tsp}$ of Figure~\ref{fig:tsp}. In this figure, the ports
linked to data (use and use-multiple ports of SYNC and K) are
represented but are not connected. Moreover, each computation
component is an instance of a specific K component using
the identification name of the computation in the MSL file.
Thus, this figure represents what the component \texttt{Computations} of Figure~\ref{fig:mscac:assembly} contains if an hybrid parallelization is performed. On the other hand, if a data parallelization is performed, a single $SEQ$ component is used by \texttt{Time}, which itself uses in the appropriate order the computation kernels and synchronizations. Among those kernels, a single one \emph{fusionned} represents $c_3$ and $c_5$.

>>>>

%----------------------------------------
%\subsection{Overview}
%----------------------------------------
This section describes the projection of a MSL program to a component-based runtime, including the components used to implement the static scheduling of $\Gamma_{tsp}$.
The section first gives an overview of component models, and especially the Low Level Components (\llc) used in this work.

@@

Many component models exist, each of them with its own specificities.
Well known component models include, for example, the CORBA Component Model (CCM)~\cite{corba:omg06}, and the Grid Component Model (GCM)~\cite{Baude} for distributed computing, while the Common Component Architecture (CCA)~\cite{Bernholdt01052006}, and Low Level Components (\llc)~\cite{l2c} are HPC-oriented.
This work makes use of \llc for the experiments.
%Let therefore introduce this model in more details.

@@

\llc~\cite{l2c} is a minimalist \texttt{C++} based HPC-oriented component model
where a component extends the concept of class.
The services offered by the components are specified trough $provide$ ports,
those used either by $use$ ports for a single service instance,
or $use-multiple$ ports for multiple service instances.
Services are specified as \texttt{C++} interfaces.
\llc also offers $MPI$ ports that enable components to share MPI communicators.
Components can also have attribute ports to be configured.
%
As illustrated in Figures~\ref{fig:ports}, a $provide$ port is
represented by a white circle, a $use$
port with a black circle, a $use-multiple$ port by a black circle with
a white $m$ in it. MPI port are
connected with a black rectangle.
A \llc-based application is a static \emph{assembly} of components instances and connections between their ports.
Such an assembly is described in LAD, an XML dialect.

@@

%----------------------------------------
\subsection{MSP Component Runtime Overview}
%----------------------------------------

@@

  \path[-]
    (Dp) edge node {} (D)
    (D) edge node {} (Du1)
        edge node {} (Du2)
        edge node {} (Du3)
    (DDSp1) edge node {} (DDS)
    (Tp) edge node {} (T)
    (T)  edge node {} (Tu)
    (Cp) edge node {} (C)
    (C) edge node {} (Cu)
    (Datap2) edge node {} (Data)
    (Datap1) edge node {} (Data)
    (Data) edge node {} (Datau)
          edge node {} (mpidata)
    (DDSp2) edge node {} (DDS);
\end{tikzpicture}
\vspace*{.5em}
\caption{MSCAC Component-based Runtime Overview.}
\label{fig:mscac:assembly}
\end{center}
\end{figure}

As described in Section~\ref{sect:mscac}, the compiler generates four independent parts in a component assembly in addition to a static part as represented in Figure~\ref{fig:mscac:assembly}: each rounded box represents one component instance, drawn with simple lines, or more, drawn with double lines. For example, the component \texttt{Data} is instantiated more than once.
\texttt{Driver} is static and drives the whole application.
The four other parts are generated based on the content of the different sections of the MSL program: \texttt{Computations}, from the parallelization of $\Gamma$ (done by MSC as explained in Section~\ref{sect:msc}), \emph{DDS} from $\mathcal{M}$ and $\mathcal{E}$, \emph{Data} from $\Delta$ and $\mathcal{D}$, and \emph{Time} from $T$.

@@

\texttt{DDS} manages the structure and the partitioning of the mesh, a single instance is generated as MSL only supports a single mesh for now.
The back-end presented in this paper uses the distributed data structure proposed in SkelGIS library~\cite{CPE:CPE3494}.
\texttt{DDS} is used by \texttt{Data}, for which one instance represents one data element in the MSL program.
\texttt{Time} is parametrized by the \texttt{iteration} value from MSL and uses the \texttt{Computations} component as many times as required.

\texttt{Computations} contains the static scheduling $\Gamma_{tsp}$, or the data parallel transformation $\Gamma_{data}$, both computed by MSC.
Both transformations, $\Gamma_{tsp}$ and $\Gamma_{data}$, are encoded using three specific components, detailed in Section~\ref{sect:control}, which manage the $P$, $S$, and $sync$ operations.
At the leaves of $\Gamma_{tsp}$ and $\Gamma_{data}$, the computation components instances (denoted $K$), provided by the user, are connected to \texttt{Data}.
As shown in Figure~\ref{fig:k}, a $K$ component provides a port to launch the computation and exposes a use port for each data element manipulated in the computation; a star is added on the use port to \texttt{Data} to denote it.
\begin{figure}
\begin{center}
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[component] (k) at (0,0) {$K$};
   \node[provide] (p) at (-1.5,0) {};
   \node[use] (u) at (1.5,0) {};
   \node[right=0.2 of u] (star) {*};

  \path[-]
    (p) edge node {} (k)
    (k) edge node {} (u);
\end{tikzpicture}
\end{center}
\caption{A kernel component $K$.}
\label{fig:k}
\end{figure}

A complete assembly is obtained by replicating the component assembly of Figure~\ref{fig:mscac:assembly} on available processors (or cores). Connections is done with MPI ports of \texttt{Data} components.

As a result of this approach, each part of the generated code is rather independent.
Changing the main time loop to use a convergence criteria rather than a fixed number of iterations would only require changing the \texttt{Time} component.
Similarly, changing the approach used for scheduling would only impact the code generated in \texttt{Computations}.
The technology used for data parallelism can also be changed by replacing the \texttt{DDS} and \texttt{Data} components, and by using the adequate interfaces in $K$. As a matter of fact, we have started evaluating an alternative based on Global Arrays~\cite{Nieplocha:2006:AAP:1125980.1125985} instead of SkelGIS.

%----------------------------------------
\subsection{Control Components and Dump}
\label{sect:control}
%----------------------------------------
The series-parallel tree decomposition $\Gamma_{tsp}$ represents a static scheduling of the simulation. $\Gamma_{data}$, on the other hand, is an optimized merged ordered list of computations. Both are dumped to components by generating an assembly that exactly matches their structure.
We introduce what we call \emph{control components} to represent all nodes of $\Gamma_{tsp}$ or $\Gamma_{data}$.
These components can be used for any simulation, which increases code reuse between simulations.
A control component exposes a single \emph{provide} port containing a \emph{control-only} method, without any parameter.
Three control components have been needed: SEQ, PAR, and SYNC. They are represented in Figure~\ref{fig:ctrlcomponents}.

\item[Synchronization component (SYNC)] It is the direct representation to an update leaf of $\Gamma_{tsp}$ or $\Gamma_{data}$. This component calls the synchronization of a given data. It exposes a \emph{use} port to be connected to the data to be updated.
\end{description}

Then, from $\Gamma_{tsp}$ or $\Gamma_{data}$, and using the control components described above,
a direct dump can be done to a component assembly for a data or hybrid (data
and task) parallelization of a simulation.
Figure~\ref{fig:tsp:assembly} displays the assembly part corresponding to
$\Gamma_{tsp}$ of Figure~\ref{fig:tsp}. In this figure, the ports
linked to data (use and use-multiple ports of SYNC and K) are
represented but are not connected for readability. Moreover, each computation
component is an instance of a specific K component using
the identification name of the computation of an MSL instance.
Thus, this figure represents what the component \texttt{Computations} of Figure~\ref{fig:mscac:assembly} contains if an hybrid parallelization is performed. On the other hand, if a data parallelization is performed, a single $SEQ$ component is used by \texttt{Time}, which invokes in the appropriate order the computation kernels and synchronizations. Among those kernels, a single \emph{merged} kernel would represent $c_3$ and $c_5$.

@@

  \path[-]
    (p) edge node {} (sync)
    (sync) edge node {} (u);
\end{tikzpicture}
\\
a) Comp. SEQ
\hspace{\fill}
b) Comp. PAR
\hspace{\fill}
c) Comp. SYNC
\\
\caption{The three control components.}
\label{fig:ctrlcomponents}
\end{figure}
